---
title: "Sample code to train maxent habitat model for a coral species"
output: html_notebook
---

This notebook and related dataset has been produced to support the publication of the article "Paris Agreement could prevent regional mass extinctions of coral species", by authors Elena Couce, Benjamin Cowburn, David Clare, Joanna K. Bluemel, published on Global Change Biology, 2023. For details on the modelling please see the publication.


Installing and loading required packages:
```{r warning= FALSE}
if(!("sp" %in% rownames(installed.packages()))) 
  install.packages("sp")
if(!("sf" %in% rownames(installed.packages()))) 
  install.packages("sf")
if(!("raster" %in% rownames(installed.packages()))) 
  install.packages("raster")
if(!("maptools" %in% rownames(installed.packages()))) 
  install.packages("maptools")
if(!("dismo" %in% rownames(installed.packages()))) 
  install.packages("dismo")
if(!("rJava" %in% rownames(installed.packages()))) 
  install.packages("rJava")
library(sp)
library(sf)
library(raster)
library(maptools)
library(dismo)
library(rJava)
```
About the use of MaxEnt in the package "dismo":
In order to train MaxEnt models using the R package "dismo", it needs access to the MaxEnt software. 
This is a java program that can be downloaded from: https://biodiversityinformatics.amnh.org/open_source/maxent/
The file 'maxent.jar' (MaxEnt version 3.3.3b or higher) needs to be placed in the 'java' folder of the "dismo" package,
which is the one returned by:
```{r}
system.file("java", package="dismo")
```


Loading relevant data for the analysis:
```{r}
data_dir = 'your-directory-here'
load(paste0(data_dir,'/analysis_data_coral_species.RData'))
load(paste0(data_dir,'/Environmental_1.RData'))
load(paste0(data_dir,'/Environmental_2.RData'))
ENV_DATA = ENV_DATA1
ENV_DATA@data = cbind(ENV_DATA@data, ENV_DATA2@data)
rm(ENV_DATA1, ENV_DATA2)
```
Data that has just been loaded:
- "ENV_DATA": 'SpatialGridDataFrame' object (from library "sp") that contains all the gridded environmental data used for the models
- "Reefs": 'SpatialPointsDataFrame' object (from library "sp") that contains the reef location data (given as the centers of all the grid cells with at least 1 reef in them). The 'ID' column relates it to the 'cell_id' column in ENV_DATA
- "world": 'SpatialPolygons' object (from library "sp") with the outline of the world's landmass, used for the plots. 
- "connectivity_matrix": Matrix with the connectivity data of all grid cells to each other, from Wood et al, 2014. The names of rows and columns refer to the 'cell_id' values in "ENV_DATA". 
- "IPcoralrangeboundaries_Horastrea.indica": A 'SpatialPolygonsDataFrame' object (from library "sp") with the range for coral species "Horastrea indica", which has been chosen for this example. This data is from Hughes, Connolly and Keith, 2013.



Plotting all reef cells (map centered in the Pacific):
```{r warning= FALSE}
plot(world,col='lightsteelblue',border='lightsteelblue') 
plot(Reefs,add=T,pch='.',cex=0.8,warnings=F)
```


Plotting range for the chosen coral species "Horastrea indica".
Data downloaded from Hughes, Connolly and Keith, 2013, Ecology 94(7)
```{r warning= FALSE}
plot(world,col='lightsteelblue',border='lightsteelblue', main='Range for Horastrea indica (from Hughes et al, 2013)') 
plot(Reefs,add=T,pch='.',cex=0.8)
plot(IPcoralrangeboundaries_Horastrea.indica,add=T,col='transparent',border='red',lwd=2)
```


Reef cells within range (i.e., "Presence sites") and connected reef cells, 
using connectivity data from: Wood, Paris, Ridgwell & Hendy, 2014, Global Ecology and Biogeography 23(1):
```{r warning= FALSE}
# Presences:
reefs_in_range = over(Reefs,IPcoralrangeboundaries_Horastrea.indica)$range
cells_species_present = Reefs$ID[!is.na(reefs_in_range)]

# Connected reef cells:
connected_cells = colnames(connectivity_matrix)[colSums(connectivity_matrix[rownames(connectivity_matrix) %in% cells_species_present,])>0]
connected_cells = connected_cells[!connected_cells %in% cells_species_present] 
```
  
  
Plotting presence and connected cells:
```{r warning= FALSE}
par(oma=c(0,0,0, 6))
plot(world,col='lightsteelblue',border='lightsteelblue',main='Horastrea indica') 
plot(Reefs,pch=16,col='blue',cex=0.3,add=T) # Plotting all reefs
plot(Reefs[Reefs$ID %in% cells_species_present,],pch=16,col='gold',add=T,cex=0.3) # Adding Presence cells
plot(Reefs[Reefs$ID %in% connected_cells,],pch=16,col='red',add=T,cex=0.3) # Adding Connected cells
plot(IPcoralrangeboundaries_Horastrea.indica,add=T,col='transparent',border='black',lwd=1)
par(fig = c(0, 1, 0, 1), oma = c(0,0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE, xpd=T)
plot.new()
legend('right',col=c('black','gold','red','blue','powderblue'),pch=c(NA,16, 16, 16),
       pt.cex=2, lty = c(1, NA, NA, NA), lwd=3,
       legend = c('Species range', 'Presence reefs', 'Connected reefs', 'Other reefs'), cex=1)
```

Defining study area for the species: 
```{r warning= FALSE}
# Changing to a projected reference system (world cylindrical equal area) for the "Reefs" sites:
Reefs_proj = spTransform(Reefs, CRS("+proj=cea +lon_0=0 +lat_ts=0 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs"))

all_cells_in_study_area = c(cells_species_present,connected_cells)
buffer_radius=200000 #200 km for spawning species
points = st_as_sf(as(Reefs_proj[Reefs_proj$ID %in% all_cells_in_study_area,],'SpatialPoints'))
circles_pr <- st_buffer(points, buffer_radius)
circles_pr <- st_wrap_dateline(st_transform(circles_pr, crs=proj4string(Reefs)), options = c("WRAPDATELINE=YES"))
circles_pr = as(circles_pr, "Spatial")
study_area = unionSpatialPolygons(circles_pr,IDs = rep(1,times=length(circles_pr)))
proj4string(study_area)= proj4string(Reefs)
```


Plotting study area, shallow water mask, and presence/connected reef sites:
```{r warning= FALSE}
par(oma=c(0,0,0,0))
plot(study_area,main='Horastrea indica',col='white') # to set appropriate plot boundaries
image(ENV_DATA['Shallow_mask'],col=c('transparent','powderblue'),add=T)
plot(study_area,add=T)
plot(world,col='lightsteelblue',border='lightsteelblue',add=T) 
plot(Reefs[Reefs$ID %in% cells_species_present,],pch=16,col='gold',add=T,cex=0.3) # Adding Presence cells
plot(Reefs[Reefs$ID %in% connected_cells,],pch=16,col='red',add=T,cex=0.3) # Adding Connected cells
#par(fig = c(0, 1, 0, 1), oma = c(0,0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE, xpd=T)
#plot.new()
legend('bottomright',col=c('black','gold','red','powderblue'),pch=c(NA,16,16,NA),
       pt.cex=2, lty = c(1, NA, NA, NA), lwd=3,fill = c(NA, NA, NA, 'powderblue'), border = c(NA, NA, NA, 'powderblue'),
       legend = c('Study Area', 'Presence reefs', 'Connected reefs', 'Shallow water mask'), cex=1)
```


Writing the presence/absences and the environmental data used to train the MaxEnt models.
Data will be saved in the same directory as the previously loaded data is in,
in a subdirectory called "Horastrea.indica":
```{r warning= FALSE}
species = "Horastrea.indica"

# Creating output folder for the species:
dir.create(species)

# Restricting to study area:
env_data_species = ENV_DATA[study_area,]

# Applying shallow water mask:
env_data_species@data[is.na(env_data_species$Shallow_mask) | !env_data_species$Shallow_mask,] = NA

# Writing distribution data:
Presences_coords = coordinates(ENV_DATA)[which(ENV_DATA$cell_id %in% cells_species_present),]
distribution_data=data.frame(ID=species,X=Presences_coords[,1],Y=Presences_coords[,2])
write.csv(distribution_data, file = paste0(data_dir,'/',species,'/Presences.csv'), row.names = FALSE)

# Writing environmental data:
Env_var2use <- c("avg_depth", "highest_point", "K490_max", "K490_mean",
                 "K490_min", "NO3_max", "NO3_mean", "NO3_min", "O2_max", 
                 "O2_mean", "O2_min", "PAR_max", "PAR_mean", "PAR_min", 
                 "PO4_max", "PO4_mean", "PO4_min", "sal_max", "sal_mean", 
                 "sal_min","SST_mean", "SST_max", "SST_min", "SST_SD", 
                 "SST_range", "Dust", "Cyclone", "Irr_mean", "Irr_max", 
                 "Irr_min", "Z_mean", "Z_max", "Z_min", "Speed_mean", 
                 "Speed_max", "Speed_min", "UKESM_Aragonite_mean", 
                 "UKESM_sal_mean", "UKESM_SST_mean")
for(vr in Env_var2use)
  write.asciigrid(env_data_species, fname=paste0(data_dir,"/",species,"/",vr,".asc"), attr = vr)
```



Generating background points within the study region, 
and extracting environmental data for presences and background points:
```{r warning= FALSE}
# Read presences and convert to spatial:
presences <- read.csv(paste0(data_dir,"/",species,'/Presences.csv'))
presences <- SpatialPoints(presences[,c('X','Y')])

# Read environmental data and put it in SpatialGridDataFrame:
Env_data = as(stack(paste0(data_dir,"/",species,'/',Env_var2use,'.asc')),'SpatialGridDataFrame')

# 1) Missing data imputation (as the median within the study area):
Env_data$study_mask = !is.na(Env_data$SST_mean)
for(var in Env_var2use)
  Env_data[[var]][is.na(Env_data[[var]]) & Env_data$study_mask] = median(Env_data[[var]][Env_data$study_mask], na.rm = TRUE)
Env_data$study_mask=NULL
Env_data = as(Env_data,'SpatialPixelsDataFrame')

# 2) Choose randomly 10000 background points in the study region:
set.seed(3)
background = spsample(Env_data, n=10000, type='random')

# 3) Environmental data for presences and background points:
background = cbind(coordinates(background),over(background,Env_data))
names(background)[1:2]=c('X','Y')
presences = cbind(Species=species,coordinates(presences),over(presences,Env_data))
names(presences)[1:3]=c('Species','X','Y')

# 4) Getting rid of presences that fall outside study region:
presences = presences[complete.cases(presences),]

# 5) Saving presences and background data to file:
write.csv(background, row.names = F, file=paste0(data_dir,"/",species,'/Background_data.csv'))
write.csv(presences, row.names = F, file=paste0(data_dir,"/",species,'/Presences_data.csv'))
```


Training MaxEnt model using the R package "dismo":
```{r warning= FALSE}
# MaxEnt arguments:
MaxEnt_args=c("responsecurves", "betamultiplier=1", "jackknife", "replicates=1", 
              "randomtestpoints=0", "doclamp", "maximumiterations=5000")

# Load presence and background data:
data.train <- read.csv(paste0(data_dir,"/",species,"/Presences_data.csv"), header=TRUE)
backgr <- read.csv(paste0(data_dir,"/",species,"/Background_data.csv"), header=TRUE)

# Combine presences and background points in one dataframe:
x <- rbind(data.train[,Env_var2use], backgr[,Env_var2use])

# Create vector containing 1 for presences and 0 for background) values to indicate which records are which
P <- c(rep(1,times=dim(data.train)[1]), rep(0,times=dim(backgr)[1]))

# Training MaxEnt model:
model_output_folder =  paste0(data_dir,"/",species,'/MaxEnt_model_',species,'_clamping')
dir.create(model_output_folder)
maxent_mod <- maxent(x=x, p=P, args=MaxEnt_args, removeDuplicates=TRUE, path=model_output_folder)
```

Plotting all environmental variables used for model training (circa 1990s):
```{r}
env_training = stack(ENV_DATA[Env_var2use])
plot(subset(env_training, Env_var2use[1:13], drop=TRUE))
plot(subset(env_training, Env_var2use[14:26], drop=TRUE))
plot(subset(env_training, Env_var2use[27:39], drop=TRUE))
```

Using the MaxEnt model to predict to training conditions:
Note: At the time the work was carried out, the package "dismo" did not implement a way to use clamping in the predictions of a MaxEnt model. Therefore we used a workaround, calling MaxEnt directly from the windows terminal with all relevant arguments. This part of the code will likely need to be modified for operating systems other than windows.
```{r}
# Creating folder for the output:
pred_folder = paste0(model_output_folder,'/prediction_training')
dir.create(pred_folder)

# Since Horastrea indica is not one of the species we have classified as highly susceptible to bleaching, we project using clamping and constant-response

# Writing environmental data used by MaxEnt for the prediction:
writeRaster(env_training, filename=paste0(pred_folder, "\\", names(env_training)), format="ascii", bylayer=T, crs=T, overwrite=TRUE)

# Maxent arguments for the prediction:
output_file <- paste0(pred_folder, "\\Projection_training.asc") 
ARGS_Clamp <- c(" doclamp=true outputformat=logistic writeclampgrid=true fadebyclamping=false") # arguments for the prediction
lambdas <- paste0(model_output_folder,'/species.lambdas') # Maxent model to use

# Path to MaxEnt java file 'maxent.jar':
maxent <- paste(system.file(package="dismo"), "/java/maxent.jar", sep='')

# System call to the MaxEnt java program:
invisible(system(paste0("java -mx1000m -cp \"", maxent, "\" density.Project \"", lambdas, "\" \"", pred_folder, "\" \"", output_file, "\" ", ARGS_Clamp)))
```


Plotting prediction and species study region:
```{r}
suitability  <- readAll(raster(output_file, native=F))
par(oma=c(0,0,0,0))
plot(study_area,main='Suitability for Horastrea indica, 1990s',col='white') # to set appropriate plot boundaries
plot(suitability,add=T)
plot(study_area,add=T)
plot(world,col='lightsteelblue',border='lightsteelblue',add=T) 
```
    


